Pre working group data exploration: Mainly wrt coral cover

### 2024-08-15 julianna_algalMetric_wrangling.Rmd
Updated wrangling code to fix sand bug for algal cover of hard substrate

# metadata matrix
algae %>% 
  # bring in sand data
  full_join(sand) %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg]), # percent cover of algae
         AlgCov_percHard = case_when(
           sand == 0 ~ AlgCov_perc, # since we can't divide by a zero
           TRUE ~ AlgCov_perc/sand 
           )) %>% # percent cover of hard substrate
   # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, "_", hab_year)) %>% 
  select(year, location, site, habitat, island_side, site_habitat, 
         hab_year, island_side_hab_year, transect, quadrat, AlgCov_perc, 
         AlgCov_percHard, sand, coral)  -> sppMeta
```

@@

```{r look for zeros}
sppMat %>% 
  mutate(AlgCov_perc = rowSums(across(where(is.numeric)))) %>% 
  filter(AlgCov_perc == 0) %>% dim() # 672 with no algae at all
```

@@

# spp matrix where 0 observations are filtered out
algae %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgCov_perc != 0) %>% 
  select(-year, -location, -island_side, -site, -habitat, -site_habitat, -transect, -quadrat, -AlgCov_perc) -> sppMat0
```

```{r metadata matix}
# matching metadata matrix
algae %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgCov_perc != 0) %>%
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, hab_year)) -> sppMeta0
```

@@

# spp matrix
algae %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgCov_perc != 0) %>% 
  filter(habitat == "fringing") %>% 
  select(-year, -location, -island_side, -site, -habitat, -site_habitat, -transect, -quadrat, -AlgCov_perc) -> sppMat0fring

# metadata matrix
algae %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgCov_perc != 0) %>%
   filter(habitat == "fringing") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, hab_year),
         island_side_year = paste0(island_side, year)) -> sppMeta0fring
```

@@

# spp matrix
algae %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgCov_perc != 0) %>% 
  filter(habitat == "backreef") %>% 
  select(-year, -location, -island_side, -site, -habitat, -site_habitat, -transect, -quadrat, -AlgCov_perc) -> sppMat0back

# metadata matrix
algae %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgCov_perc != 0) %>%
   filter(habitat == "backreef") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, hab_year),
         island_side_year = paste0(island_side, year)) -> sppMeta0back
```

@@

# spp matrix
algae %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgCov_perc != 0) %>% 
  filter(habitat == "outer_10") %>% 
  select(-year, -location, -island_side, -site, -habitat, -site_habitat, -transect, -quadrat, -AlgCov_perc) -> sppMat0outer10

# metadata matrix
algae %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgCov_perc != 0) %>%
   filter(habitat == "outer_10") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, hab_year),
         island_side_year = paste0(island_side, year)) -> sppMeta0outer10
```

@@

# spp matrix
algae %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgCov_perc != 0) %>% 
  filter(habitat == "outer_17") %>% 
  select(-year, -location, -island_side, -site, -habitat, -site_habitat, -transect, -quadrat, -AlgCov_perc) -> sppMat0outer17

# metadata matrix
algae %>% 
  mutate(AlgCov_perc = rowSums(algae[startAlg:endAlg])) %>% 
  filter(AlgCov_perc != 0) %>%
   filter(habitat == "outer_17") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, hab_year),
         island_side_year = paste0(island_side, year)) -> sppMeta0outer17
```

@@

```{r join}
algDiv %>% 
  full_join(betaDHabYr) %>% 
  # follow a cleaner naming convention to match Noam's
  rename(algae_cover = AlgCov_perc,
         algae_cover_hard = AlgCov_percHard,
         algae_richness = Richness,
         algae_shannon = Shannon) %>% 
  # get just the columns we need
  select(-site_habitat, -hab_year, -island_side_hab_year) -> quadData

@@

```{r get averages}
quadData %>% 
  group_by(year, site, habitat, island_side) %>% 
  summarize(coral_cover_mean = mean(coral),
            N = n(),
            coral_cover_se = sd(coral)/sqrt(N),
            algae_cover_mean = mean(algae_cover),
            algae_cover_se = sd(algae_cover)/sqrt(N),
            algae_cover_hard_mean = mean(algae_cover_hard),
            algae_cover_hard_se = sd(algae_cover_hard)/sqrt(N),
            algae_richness_mean = mean(algae_richness),
            algae_richness_se = sd(algae_richness)/sqrt(N),
            algae_shannon_mean = mean(algae_shannon),
            algae_shannon_se = sd(algae_shannon)/sqrt(N),
            centroid_dist_habyear_mean = mean(hab_yr_distcentroid, na.rm = TRUE)
            ) -> siteData

@@

Apply it when we summarize

```{r calculate CV and summarize}
quadData %>% 
  # don't include year this time
  group_by(site, habitat, island_side) %>%
  summarize(
    coral_cover_mean = mean(coral),
    algae_richness_mean = mean(algae_richness), 
    algae_richness_se = sd(algae_richness)/sqrt(n()),
    algae_richness_cv = CV(algae_richness),
    algae_shannon_mean = mean(algae_shannon), 
    algae_shannon_se = sd(algae_shannon)/sqrt(n()),
    algae_shannon_cv = CV(algae_shannon),
    # for total algal cover
    algae_cover_mean = mean(algae_cover),
    algae_cover_se = sd(algae_cover)/sqrt(n()),
    algae_cover_cv = CV(algae_cover),
    # and for % of hard substrate
    algae_cover_hard_mean = mean(algae_cover_hard),
    algae_cover_hard_se = sd(algae_cover_hard)/sqrt(n()),
    algae_cover_hard_cv = CV(algae_cover_hard),
    ) -> alltimeCV
```

>>>>

# metadata matrix
algae %>% 
  # bring in sand data
  full_join(sand) %>% 
  mutate(algae_cover = rowSums(algae[startAlg:endAlg]), # percent cover of algae
         algae_cover_hard = case_when(
           sand == 100 ~ NA, # since we can't divide by a zero
           TRUE ~ (algae_cover/(100-sand))*100 
           )) %>% # percent cover of hard substrate
   # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, "_", hab_year)) %>% 
  select(year, location, site, habitat, island_side, site_habitat, 
         hab_year, island_side_hab_year, transect, quadrat, algae_cover, 
         algae_cover_hard, sand, coral)  -> sppMeta
```

@@

```{r look for zeros}
sppMat %>% 
  mutate(algae_cover = rowSums(across(where(is.numeric)))) %>% 
  filter(algae_cover == 0) %>% dim() # 672 with no algae at all
```

@@

# spp matrix where 0 observations are filtered out
algae %>% 
  mutate(algae_cover = rowSums(algae[startAlg:endAlg])) %>% 
  filter(algae_cover != 0) %>% 
  select(-year, -location, -island_side, -site, -habitat, -site_habitat, -transect, -quadrat, -algae_cover) -> sppMat0
```

```{r metadata matix}
# matching metadata matrix
algae %>% 
  mutate(algae_cover = rowSums(algae[startAlg:endAlg])) %>% 
  filter(algae_cover != 0) %>%
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, hab_year)) -> sppMeta0
```

@@

# spp matrix
algae %>% 
  mutate(algae_cover = rowSums(algae[startAlg:endAlg])) %>% 
  filter(algae_cover != 0) %>% 
  filter(habitat == "fringing") %>% 
  select(-year, -location, -island_side, -site, -habitat, -site_habitat, -transect, -quadrat, -algae_cover) -> sppMat0fring

# metadata matrix
algae %>% 
  mutate(algae_cover = rowSums(algae[startAlg:endAlg])) %>% 
  filter(algae_cover != 0) %>%
   filter(habitat == "fringing") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, hab_year),
         island_side_year = paste0(island_side, year)) -> sppMeta0fring
```

@@

# spp matrix
algae %>% 
  mutate(algae_cover = rowSums(algae[startAlg:endAlg])) %>% 
  filter(algae_cover != 0) %>% 
  filter(habitat == "backreef") %>% 
  select(-year, -location, -island_side, -site, -habitat, -site_habitat, -transect, -quadrat, -algae_cover) -> sppMat0back

# metadata matrix
algae %>% 
  mutate(algae_cover = rowSums(algae[startAlg:endAlg])) %>% 
  filter(algae_cover != 0) %>%
   filter(habitat == "backreef") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, hab_year),
         island_side_year = paste0(island_side, year)) -> sppMeta0back
```

@@

# spp matrix
algae %>% 
  mutate(algae_cover = rowSums(algae[startAlg:endAlg])) %>% 
  filter(algae_cover != 0) %>% 
  filter(habitat == "outer_10") %>% 
  select(-year, -location, -island_side, -site, -habitat, -site_habitat, -transect, -quadrat, -algae_cover) -> sppMat0outer10

# metadata matrix
algae %>% 
  mutate(algae_cover = rowSums(algae[startAlg:endAlg])) %>% 
  filter(algae_cover != 0) %>%
   filter(habitat == "outer_10") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, hab_year),
         island_side_year = paste0(island_side, year)) -> sppMeta0outer10
```

@@

# spp matrix
algae %>% 
  mutate(algae_cover = rowSums(algae[startAlg:endAlg])) %>% 
  filter(algae_cover != 0) %>% 
  filter(habitat == "outer_17") %>% 
  select(-year, -location, -island_side, -site, -habitat, -site_habitat, -transect, -quadrat, -algae_cover) -> sppMat0outer17

# metadata matrix
algae %>% 
  mutate(algae_cover = rowSums(algae[startAlg:endAlg])) %>% 
  filter(algae_cover != 0) %>%
   filter(habitat == "outer_17") %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, island_side) %>% 
  # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         island_side_hab_year = paste0(island_side, hab_year),
         island_side_year = paste0(island_side, year)) -> sppMeta0outer17
```

@@

```{r join}
algDiv %>% 
  full_join(betaDHabYr) %>% 
  # follow a cleaner naming convention to match Noam's
  rename(algae_richness = Richness,
         algae_shannon = Shannon) %>% 
  # get just the columns we need
  select(-site_habitat, -hab_year, -island_side_hab_year) -> quadData

@@

```{r get averages}
quadData %>% 
  group_by(year, site, habitat, island_side) %>% 
  summarize(coral_cover_mean = mean(coral),
            N = n(),
            coral_cover_se = sd(coral)/sqrt(N),
            algae_cover_mean = mean(algae_cover),
            algae_cover_se = sd(algae_cover)/sqrt(N),
            algae_cover_hard_mean = mean(algae_cover_hard, na.rm = TRUE),
            algae_cover_hard_se = sd(algae_cover_hard, na.rm = TRUE)/sqrt(N),
            algae_richness_mean = mean(algae_richness),
            algae_richness_se = sd(algae_richness)/sqrt(N),
            algae_shannon_mean = mean(algae_shannon),
            algae_shannon_se = sd(algae_shannon)/sqrt(N),
            centroid_dist_habyear_mean = mean(hab_yr_distcentroid, na.rm = TRUE)
            ) -> siteData

@@

```{r calculate CV and summarize}
quadData %>% 
  # don't include year this time
  group_by(site, habitat, island_side) %>%
  summarize(
    coral_cover_mean = mean(coral),
    algae_richness_mean = mean(algae_richness), 
    algae_richness_se = sd(algae_richness)/sqrt(n()),
    algae_richness_cv = CV(algae_richness),
    algae_shannon_mean = mean(algae_shannon), 
    algae_shannon_se = sd(algae_shannon)/sqrt(n()),
    algae_shannon_cv = CV(algae_shannon),
    # for total algal cover
    algae_cover_mean = mean(algae_cover),
    algae_cover_se = sd(algae_cover)/sqrt(n()),
    algae_cover_cv = CV(algae_cover),
    # and for % of hard substrate
    algae_cover_hard_mean = mean(algae_cover_hard, na.rm = TRUE),
    algae_cover_hard_se = sd(algae_cover_hard, na.rm = TRUE)/sqrt(n()),
    algae_cover_hard_cv = CV(algae_cover_hard),
    ) -> alltimeCV
```

### 2024-08-15 julianna_coral_macroalg_wrangling.Rmd
Updated wrangling code to fix sand bug for algal cover of hard substrate

>>>>

# Bring in the data

```{r}
# algal cover 
macroalgae <- read_csv(here("data", "algaeonly_cleaned_colnames.csv")) %>% 
  mutate(island_side = case_when(site == "lter_1" ~ "north",
                                 site == "lter_2" ~ "north",
                                 site == "lter_3" ~ "east",
                                 site == "lter_4" ~ "east",
                                 site == "lter_5" ~ "west",
                                 site == "lter_6" ~ "west",
                                 TRUE ~ "UdINGLEbUTT")) %>% 
  select(-algal_turf, -crustose_corallines, -cyanophyta, -damselfish_turf, -lithophyllum_kotschyanum,
         -neogoniolithon_brassica_florida, -sporolithon_sp) # got this from the cleaning code--should Peysonellia be included?

# sand cover and coral cover
sand_coral <- read_csv(here("data", "allbenthicdata_cleaned_colnames.csv")) %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, sand, coral)
```

# Calculate macroalgal alpha diversity

## Make matrices

Make a spp matrix to use with `vegan`

```{r make species matrix}
# matrix of just species and their relative percent covers in the same order as the algae tibble
macroalgae %>% 
  select(-year, -location, -site, -island_side, -habitat, -site_habitat, -transect, -quadrat) -> sppMat
```

### 2024-08-15 julianna_data_exploration.Rmd
Updated wrangling code to fix sand bug for algal cover of hard substrate

# load packages
require(here) # relative file paths
require(tidyverse) # data wrangling
require(vegan) # community/diversity analyses
require(lme4) # glmer()
require(emmeans) # emmeans()
require(viridis) # color palette
require(MoMAColors) # another color palette
require(patchwork) # putting figures together
require(BiodiversityR)
require(ggsci) # scientific jounral and sci-fi themed color palette
```

>>>>

require(librarian)

# load packages
librarian::shelf(
  here, # relative file paths
  tidyverse,  # data wrangling
  vegan, # data wrangling
  lme4, # glmer()
  emmeans, # emmeans()
  viridis, # color palette
  MoMAColors, # another color palette: devtools::install_github("BlakeRMills/MoMAColors")
  patchwork, # putting figures together
  BiodiversityR,
  ggsci # scientific jounral and sci-fi themed color palette
)
